Gamma-Ray Burst Afterglows In The Multi-Messenger Era: Numerical Models and Closure Relations

### 2019-09-25 structuredJets.tex
Arxiv Version

The binary neutron star merger event \gwbns{}, followed quickly by the short gamma-ray burst \grbbns{}, provided a new view on gamma-ray burst (GRB) afterglows \citep{Abbott:2017ab, Abbott:2017ac}.  Unlike typical afterglows, which begin bright and decay from detection with a ~week timescale, \grbbns{}'s non-thermal emission was undetectable until the first observation of X-rays nine days after the GRB \citep{Troja:2017aa}.  At this point the afterglow began steadily increasing in brightness in at all wavelengths for ~160 days \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  Very long baseline (VLBI) radio imagery over this period identified a radio core with an apparent superluminal motion, indicating the emitting surface was moving at relativistically at an oblique angle towards the Earth \citep{Mooley:2018ab}. The emission peaked at 164 days after the burst and proceeded to sharply decay at a rate $t^{-2.2}$ commensurate with other GRB afterglows \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Troja:2019ab}.

@@

In this paper, we provide a characterization of the afterglows of structured jets at all viewing angles, explicit closure relations, and a description of our  public numerical tool \afterglowpy{}. In Section \ref{sec:motivation} we review the general properties of structured jets and their necessity for GRB afterglow modelling.  In Section \ref{sec:numerical} we develop our theoretical framework and describe its numerical implementation in \afterglowpy{}.  Section \ref{sec:structuredJets} characterizes the general behavior of structured jet afterglow light curves, provides the closure relations with explicit viewing angle dependence, and relates structured jets to standard energy injection models.  Section \ref{sec:gw170817} demonstrates an application to \gwbns{}. Section \ref{sec:discussion} gives further discussion and Section \ref{sec:summary} gives the summary.  Detailed derivation of the closure relations is provided in Appendices \ref{app:derive1} and \ref{app:derive2}.

@@

We have constructed flexible models for the electromagnetic afterglows of angularly structured relativistic jets for aligned and misaligned observers.  Through analytic approximations we can identify the basic phases of the structured afterglow light curve, determining formulae for the characteristic break time scales and closure relations.  We find the closure relations depend on a single parameter we call $g$ related to the jet structure and viewing angle.  Measurements of $g$ itself are model independent, but relating it back to physical parameters of the jet is not.

We also constructed a numerical model of the afterglows of structured jets, implemented in the public \python{} package \afterglowpy{}. \afterglowpy{} computes afterglow light curves on the fly utilizing semi-analytic approximations to the jet evolution and synchrotron emission, taking into account relativistic beaming, the equal time of arrival surface, jet angular structure, trans-relativistic evolution, and jet spreading.  It is fast enough to be incorporated in MCMC parameter estimation routines.  Fitting the \gwbns{} afterglow with multiple \afterglowpy{} models allows jet parameters to be inferred, but the degeneracy between jet structure and viewing angle makes determining a particular jet profile difficult.  Incorporating data beyond the afterglow light curve, including constraints from gravitational wave observations and detailed electromagnetic followup such as VLBI, will be key to learning as much as we can from future events.

@@

>>>>

The binary neutron star merger event \gwbns{}, followed quickly by the short gamma-ray burst \grbbns{}, provided a new view on gamma-ray burst (GRB) afterglows \citep{Abbott:2017ab, Abbott:2017ac}.  Unlike typical afterglows, which begin bright and decay from detection with a ~week timescale, \grbbns{}'s non-thermal emission was undetectable until the first observation of X-rays nine days after the GRB \citep{Troja:2017aa}.  At this point the afterglow began steadily increasing in brightness at all wavelengths for $\sim160$ days \citep{Haggard:2017aa, Hallinan:2017aa, DAvanzo:2018aa,  Lyman:2018aa, Margutti:2018aa, Mooley:2018aa, Troja:2018aa, Troja:2019ab}.  Very long baseline (VLBI) radio imagery over this period identified a radio core with an apparent superluminal motion, indicating the emitting surface was moving at relativistically at an oblique angle towards the Earth \citep{Mooley:2018ab}. The emission peaked at 164 days after the burst and proceeded to sharply decay at a rate $t^{-2.2}$ commensurate with other GRB afterglows \citep{Alexander:2018aa, Fong:2019aa, Lamb:2019aa, Troja:2019ab}.

@@

In this paper, we provide a characterization of the afterglows of structured jets at all viewing angles, explicit closure relations, and a description of our  public numerical tool \afterglowpy{}. In Section \ref{sec:motivation} we review the general properties of structured jets and their necessity for GRB afterglow modelling.  In Section \ref{sec:numerical} we develop our theoretical framework and describe its numerical implementation in \afterglowpy{}.  Section \ref{sec:structuredJets} characterizes the general behavior of structured jet afterglow light curves, provides the closure relations and jet break times with explicit viewing angle dependence, and relates structured jets to standard energy injection models.  Section \ref{sec:gw170817} demonstrates an application to \gwbns{}. Section \ref{sec:discussion} gives further discussion and Section \ref{sec:summary} gives the summary.  Detailed derivation of the closure relations is provided in Appendices \ref{app:derive1} and \ref{app:derive2}.

@@

We have constructed flexible models for the electromagnetic afterglows of angularly structured relativistic jets for aligned and misaligned observers.  Through analytic approximations we can identify the basic phases of the structured afterglow light curve, determining formulae for the characteristic break time scales and closure relations.  We find the closure relations depend on a single dimensionless parameter $g$ related to the jet structure and viewing angle.  Measurements of $g$ itself are model independent, but relating it back to physical parameters of the jet is not.

We also constructed a numerical model of the afterglows of structured jets, implemented in the public \python{} package \afterglowpy{}. \afterglowpy{} computes afterglow light curves on the fly utilizing semi-analytic approximations to the jet evolution and synchrotron emission, taking into account relativistic beaming, the equal time of arrival surface, jet angular structure, trans-relativistic evolution, and jet spreading.  It is fast enough to be incorporated in MCMC parameter estimation routines, available via {\tt PyPI} and {\tt pip}, and has source code available at: \url{https://github.com/geoffryan/afterglowpy}. Fitting the \gwbns{} afterglow with multiple \afterglowpy{} models allows jet parameters to be inferred, but the degeneracy between jet structure and viewing angle makes determining a particular jet profile difficult.  Incorporating data beyond the afterglow light curve, including constraints from gravitational wave observations and detailed electromagnetic followup such as VLBI, will be key to learning as much as we can from future events.

@@

The marginalized posterior limits on the afterglow parameters presented in Table \ref{tab:MCMC} are a useful summary but do not show how the inferred parameter values correlate with each other.  Figures \ref{fig:cornerGaussian} and \ref{fig:cornerPowerlaw} show all pairwise correlations between parameters in our fits to \grbbns{} for the Gaussian and power law jet models respectively, as well as the marginalized distributions for each parameter.
